---
title: "Problem Set Rmarkdown and GGplot"
author: "Jordan Lee"
date: "2023-02-19"
output: html_document
---

## 1 Load and check data (5pt)
Your first task is to do a very simple data check:

1. (1pt) For solving the problems, and answering the questions, create a new r markdown document with a appropriate title
2. (2pt) Load data. How many rows/columns do we have?
```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("../../data/gapminder.csv")
```
The _gapminder_ data contains `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.

3. (2pt) Print a small sample of data. Does it look OK?
```{r}
gapminder %>%
  sample_n(5) %>%
  knitr::kable()
```
The sample looks good !!

## 2 Descriptive satistics (15pt)
1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

The _gapminder_ data has 3 variables representing countries _iso3, iso2,_ and _name_ 

_iso3_ has `r length(unique(gapminder$iso3))` unique values

_iso2_ has `r length(unique(gapminder$iso2))` unique values

_name_ has `r length(unique(gapminder$name))` unique values

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
```{r}
gapminder %>%
  group_by(iso2) %>%
  summarise(num = n_distinct(name)) %>%
  filter(num > 1)
```
Yes, from the code segment output above we see that the only _iso-2_ code that correspond
to more than one _name_ is 'NA' (or 'missing'). This means that there are two different
country names with the _iso-2_ field missing. Other than this, all other _iso-2_ codes
correspond to only 1 country name. 

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
```{r}
gapminder %>%
  group_by(name) %>%
  summarise(num = n_distinct(iso3)) %>%
  filter(num > 1)
```
Yes, from the code segment output above we see that the only country name that has more than one
_iso_3_ code is 'NA' (or missing). From this we can tell that there are 4 'missing' values 
in the name field, however, each of them still have/are assigned a different _iso-3_ codes.

Hint: two of these entitites are CHANISL and NLD CURACAO.

3. (2pt) What is the minimum and maximum year in these data?
```{r}
gapminder %>%
  select(time) %>%
  filter(!is.na(time)) %>%
  summarise(max_year = max(time), min_year = min(time))
```
## 3 CO2 Emmisions
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
gapminder %>%
  group_by(time) %>%
  select(time, co2, co2_PC) %>%
  filter(!is.na(time)) %>%
  summarise(co2_na = sum(is.na(co2)), co2_PC_na = sum(is.na(co2_PC))) %>%
  mutate(sum = co2_na + co2_PC_na) %>%
  arrange(desc(sum))
```
Years 2017, 2018, 2019 have the most missing _co2_ and _co2_PC_ fields

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r, message=FALSE}
gapminder %>%
  select(name, time, co2) %>%
  filter(!is.na(co2)) %>%
  filter(name %in% c("Japan", "China", "India", "United States of America", "Russian Federation")) %>%
  ggplot(aes(time, co2 / 1000000, col=name)) + 
  geom_line() +
  labs(title = "Total CO2 Emission per Year",
       x = "Time (Year)", 
       y = "CO2 Emission (Million kt)", 
       col = "Country Name")
```

The graph depicts the countries with the highest CO2 Emmisions (kt). From this,
we see that China has the highest amount of CO2 emmisions, with the U.S. coming
in second amongst the 5 countries for CO2 emmisions. After the year 2000, Russia
Japan, and India all hover around the same amount of CO2 emmisions. This graph
may indicate a positive relationship between population size and CO2 emmisions

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the
same countries. What does this figure suggest?
```{r, message=FALSE}
gapminder %>%
  select(name, time, co2_PC) %>%
  filter(!is.na(co2_PC)) %>%
  filter(name %in% c("Japan", "China", "India", "United States of America", "Russian Federation")) %>%
  ggplot(aes(time, co2_PC, col=name)) + 
  geom_line() +
  labs(title = "CO2 Emission per Capita per Year",
       x = "Time (Year)", 
       y = "CO2 Emission Per Capita (Metric Tons per Capita)", 
       col = "Country Name")
```

This graph depicts the amount of CO2 emission per person in a country. China 
actually has a lower overall CO2 emission per capita compared to other countries 
(besides India). This is interesting because when comparing this graph to the 
previous one, China has the largest overall CO2 Emission per year. This suggests
that there is a positive relationship between population size and total CO2
emission per year. 

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
Hint: Americas 2016 should be 4.80.
```{r}
gapminder %>%
  group_by(region) %>%
  select(region, co2_PC) %>%
  filter(!is.na(region)) %>%
  summarise(avg_co2_PC = mean(co2_PC, na.rm = TRUE))
```


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r, message=FALSE}
avg_co2 <- gapminder %>%
  select(region, co2_PC, time) %>%
  filter(!is.na(region), time %in% c("1960", "2016")) %>%
  group_by(region, time) %>%
  summarise(avg_co2_PC = mean(co2_PC, na.rm = TRUE))

ggplot(avg_co2, aes(region, avg_co2_PC, fill=factor(time))) +
  geom_col(position = "dodge") +
  labs(title = "Average CO2 per Capita Across Regions (Continents)",
       x = "Region (Continent Name)",
       y = "Average CO2 per Capita (Metric Ton per Capita)",
       col = "Time (Year)")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2016 for each continent? (Assume region is continent).

```{r, message=FALSE}
co2_PC_countries <- gapminder %>%
  select(region, name, co2_PC, time) %>%
  filter(time == "2016", !is.na(region), !is.na(co2_PC)) %>%
  arrange(region, desc(co2_PC)) %>%
  group_by(region)
```

Top 3 Countries from each Region
```{r}
co2_PC_countries %>%
  top_n(3, co2_PC)
```

Bottom 3 Countries from each Region
```{r}
co2_PC_countries %>%
  top_n(-3, co2_PC)
```

## 4 GDP per capita (50pt)

Let’s look at GDP per capita (GDP_PC ).

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r, message = FALSE}
gdp_PC <- gapminder %>%
  group_by(name) %>%
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>%
  filter(time == "1960", !is.na(GDP_PC), !is.na(lifeExpectancy))

ggplot(gdp_PC, aes(GDP_PC, lifeExpectancy, col = factor(region))) +
  geom_point(aes(size=totalPopulation / 1000000), alpha = 0.8) +
  labs(title = "Life Expectancy vs. GDP per Capita in 1960",
       x = "GDP per Capita",
       y = "Life Expentancy (Years)",
       col = "Region (Continents)",
       size = "Total Population (Millions)")
```


2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r, message = FALSE}
gdp_PC <- gapminder %>%
  group_by(name) %>%
  select(region, GDP_PC, lifeExpectancy, totalPopulation, time) %>%
  filter(time == "2019", !is.na(GDP_PC), !is.na(lifeExpectancy), !is.na(region))

ggplot(gdp_PC, aes(GDP_PC, lifeExpectancy, col = factor(region))) +
  geom_point(aes(size=totalPopulation / 1000000), alpha = 0.8) + 
   labs(title = "Life Expectancy vs. GDP per Capita in 2019",
       x = "GDP per Capita",
       y = "Life Expentancy (Years)",
       col = "Region (Continents)",
       size = "Total Population (Millions)")
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

For sure when looking at the two plots, there are a lot more countries in 2019
compared to in 1960. The overall GDP_PC has decreased, resulting in life expectancy
to increase in almost every region / country. This can be because of factors like
technology and access to medical support has improved through the last 60 years,
which has greatly increased the overall expectancy of life. 

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?

Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.

```{r, message=FALSE}
avg_LE <- gapminder %>%
  group_by(region, time) %>%
  select(region, time, lifeExpectancy) %>%
  filter(time %in% c("1960", "2019"), !is.na(region)) %>%
  summarise(avg_LE = mean(lifeExpectancy, na.rm = TRUE))

avg_LE
```


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.

Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up!

```{r, message=FALSE}
avg_LE %>%
  group_by(region) %>%
  mutate(prev.val = lag(avg_LE)) %>%
  filter(row_number() %% 2 == 0) %>%
  mutate(avg_growth = avg_LE - prev.val) %>%
  summarise(region, avg_growth) %>%
  arrange(desc(avg_growth))
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!

```{r}
GDP_histogram <- gapminder %>%
  filter(time == "1960" | time == "2019", !is.na(GDP_PC))
ggplot(GDP_histogram, aes(GDP_PC, fill = factor(time))) +
  geom_histogram(alpha = 0.8, position = "dodge", bins = 30) +
  labs(title = "GDP per capita for years of 1960 and 2019",
       x = "GDP per Capita",
       y = "Count",
       fill = "Year")
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)

Hint: check out the function rank()!

Hint2: 17 for 1960.

```{r}
ranks_per_year <- gapminder %>%
  select(name, lifeExpectancy, time) %>%
  filter(!is.na(lifeExpectancy), !is.na(name), time %in% c("1960", "2019")) %>%
  arrange(time, desc(lifeExpectancy)) %>%
  group_by(time) %>%
  mutate(rankings = rank(desc(lifeExpectancy)))

ranks_per_year %>%
  filter(name == "United States of America")
```


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?

Hint: 0.0904 for 1960.

```{r}
ranks_per_year %>%
  select(name, time, rankings) %>%
  group_by(time) %>%
  mutate(total_n = max(rankings)) %>%
  filter(name == "United States of America") %>%
  summarise(rr_over_ncountries = rankings/total_n)
```


## Finally tell us how many hours did you spend on this PS.

10 Hours